---
title: 'Lab 3: Panel Models'
subtitle: 'US Traffic Fatalities: 1980 - 2004'
output: 
  bookdown::pdf_document2: default
---

```{r load packages, echo=FALSE, message=FALSE}
library(GGally)
library(tibble)
library(lme4)
library(gridExtra)
library(reshape2)
library(mgcv)
library(plm)
library(lmtest)
library(aTSA)
library(urca)
library(tsibble)
library(dplyr)
library(fabletools)
library(tidyverse)
library(magrittr)
library(patchwork)
library(scales)
library(plyr)
library(tidyr)
library(ggplot2)
library(ggthemes)
library(lubridate)
library(forecast)
library(sandwich)
library(tseries)
library(vars)
library(jsonlite)
library(fable)
library(gtrendsR) 
library(zoo)
library(feasts)
library(thematic)
library(ggfortify)
library(fpp3)
require(knitr)
library(stargazer)
knitr::opts_chunk$set(tidy = TRUE, tidy.opts = list(comment = FALSE))
knitr::opts_chunk$set(comment = " ")
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```


# U.S. traffic fatalities: 1980-2004

In this lab, we are asking you to answer the following **causal** question: 

> **"Do changes in traffic laws affect traffic fatalities?"**  

To answer this question, please complete the tasks specified below using the data 
provided in `data/driving.Rdata`. This data includes 25 years of data that cover 
changes in various state drunk driving, seat belt, and speed limit laws. 

Specifically, this data set contains data for the 48 continental U.S. states from 
1980 through 2004. Various driving laws are indicated in the data set, such as the 
alcohol level at which drivers are considered legally intoxicated. There are also 
indicators for “per se” laws—where licenses can be revoked without a trial—and seat 
belt laws. A few economics and demographic variables are also included. The 
description of the each of the variables in the dataset is also provided in 
the dataset. 

```{r load data, echo = TRUE}
load(file="/Users/srilamaiti/Downloads/data/driving.RData")
glimpse(data)
desc
```


# (30 points, total) Build and Describe the Data 

1. (5 points) Load the data and produce useful features.
    - Produce a new variable, called `speed_limit` that re-encodes the data that 
    is in `sl55`, `sl65`, `sl70`, `sl75`, and `slnone`; 
    - Produce a new variable, called `year_of_observation` that re-encodes the 
    data that is in `d80`, `d81`, ... , `d04`. 
    - Produce a new variable for each of the other variables that are one-hot 
    encoded (i.e. `bac*` variable series). 
    - Rename these variables to sensible names that are legible to a reader of 
    your analysis. For example, the dependent variable as provided is called, 
    `totfatrte`. Pick something more sensible, like, `total_fatalities_rate`. 
    There are few enough of these variables to change, that you should change 
    them for all the variables in the data. (You will thank yourself later.)

# Feature Engineering

We have added state abbriviation in the data along with DC.
```{r, adding state abbriviation}
# Adding state abbriviation
state_avbbriv <- data.frame("index"=1:51,"abbr"=sort(c(state.abb,"DC")))
data <- merge(data, states, by.x = "state", by.y = "index")
```

We then added year of observation in the data set.
```{r, adding yeer of observation}
# year_of_observation
data <- data %>%
  mutate(year_of_observation = factor(year))
```

Next, we have manipulated speed limit and seat belt related columns:-
(sl55/sl65/sl70/s75/slone). Although, we have expected the data to be binary across 
speed limit columns, we have noticed fractional distributions and we devised a 
rule where max value will be considered from these percentage values. We also
created another binary column speed_limit based on speed_limit. If the value
in speed_limit is equal or more than 70, the column will be set to 1, otherwise 0.
```{r, adding speed limit and seat belt feature}
# speed_limit
data <- data %>%
  mutate(speed_limit = ifelse(sl55 >= 0.5, 55,
                       ifelse(sl65 >= 0.5, 65,
                       ifelse(sl70 >= 0.5, 70,
                       ifelse(sl75 >= 0.5, 75,
                       ifelse(slnone >= 0.5, 'none', 0)
                       ))))) 

data = data %>% mutate(
    seatbelt = factor(seatbelt), # 'seatbelt' categorizes primary or secondary
    speed_limit_70plus = ifelse(speed_limit == 55 | speed_limit == 65, 0, 1)
  )
```

```{r, blood alcohol limit}
# Rounding bac08 and bac10
data <- data %>% 
  mutate(
    bac08_processed = round(bac08),
    bac10_processed = round(bac10)
  )
# blood alcohol limit
data <- data %>% mutate(blood_alcohol_limit = ifelse(bac10_processed >= .5, .1,
                       ifelse(bac08_processed >= 0.5, .08, 0)
                       ))

```

```{r, renaming to meaningful columns}
# rename the variables to sensible names
data <- data %>%
  dplyr::rename(
    "total_fatalities_rate"                      = "totfatrte",
    "minimum_drinking_age"                       = "minage",
    "zero_tolerance_law"                         = "zerotol",
    "state_population"                           = "statepop",
    "graduated_drivers_license_law"              = "gdl",
    "per_se_laws"                                = "perse",
    "total_traffic_fatalities"                   = "totfat",
    "total_nighttime_fatalities"                 = "nghtfat",
    "total_weekend_fatalities"                   = "wkndfat",
    "total_fatalities_per_100_million_miles"     = "totfatpvm",
    "nighttime_fatalities_per_100_million_miles" = "nghtfatpvm",
    "weekend_fatalities_per_100_million_miles"   = "wkndfatpvm",
    "nighttime_fatalities_rate"                  = "nghtfatrte",
    "weekend_fatalities_rate"                    = "wkndfatrte",
    "vehicle_miles"                              = "vehicmiles",
    "unemployment_rate"                          = "unem",
    "pct_population_14_to_24"                    = "perc14_24",
    "vehicle_miles_per_capita"                   = "vehicmilespc",
    "primary_seatbelt_law"                       = "sbprim",
    "secondary_seatbelt_law"                     = "sbsecon"
  ) 

df = df[, c('year_of_observation',
            'state',
            'year',
    
            # response variables
            'total_fatalities_rate',
            'nighttime_fatalities_rate',
            'weekend_fatalities_rate',
            'total_traffic_fatalities',
            'total_nighttime_fatalities',
            'total_weekend_fatalities',
            'total_fatalities_per_100_million_miles',
            'nighttime_fatalities_per_100_million_miles',
            'weekend_fatalities_per_100_million_miles',
    
            #potential explanatory variables
            'seatbelt',
            'zero_tolerance_law',
            'graduated_drivers_license_law',
            'per_se_laws',
            'minimum_drinking_age',
            'speed_limit_70plus',
            'speed_limit',
            'primary_seatbelt_law',
            'secondary_seatbelt_law',
            'blood_alcohol_limit',
            'vehicle_miles',
            'vehicle_miles_per_capita',
    
            # econ and demographic variables
            'state_population',
            'unemployment_rate',
            'pct_population_14_to_24', 
            'vehicle_miles'
           )
]

# Check data
df %>% glimpse()
```

2. (5 points) Provide a description of the basic structure of the dataset. What is 
this data? How, where, and when is it collected? Is the data generated through a 
survey or some other method? Is the data that is presented a sample from the population, 
or is it a *census* that represents the entire population? Minimally, this should include:
    - How is the our dependent variable of interest `total_fatalities_rate` defined? 
    
> The data is for state-year-level total traffic fatalities information. It includes 
state law information that might be an indicator for traffic fatalities, like the 
speed limit, blood alcohol legal limit, seat belt laws, etc. The data was compiled 
by Donald G Freedman for the paper "Drunk living legislation and traffic fatalities:
New evidence on BAC 08 laws" - Contemporary Economic Policy 2007. In the paper 
it is noted that "Fatality data are from the Fatality Analysis Reporting System (FARS) 
compiled by NHTSA. Data on traffic legislation for the years 1982—1999 were provided 
by Thomas Dee. Earlier data on legislation were taken from Zador et al. (1989) 
and later data on legislation from the National Center for Statistics and Analysis 
at the NHTSA Web site at http://www-nrd.nhtsa.dot.gov/departments/nrd-30/ncsa/. 
Data on graduated drivers’ licenses are taken from Dee, Grabowski, and Morrisey 
(2005). State unemployment rates are from Dee and the Bureau of Labor Statistics; 
age data are from the Bureau of the Census".The outcome of interest, 
`total_fatalities_rate` is defined as the number of fatalities per 100,000 people.

3. (20 points) Conduct a very thorough EDA, which should include both graphical 
and tabular techniques, on the dataset, including both the dependent variable 
`total_fatalities_rate` and the potential explanatory variables. Minimally, this 
should include: 
    - How is the our dependent variable of interest `total_fatalities_rate` defined? 
    - What is the average of `total_fatalities_rate` in each of the years in the 
    time period covered in this dataset? 

As with every EDA this semester, the goal of this EDA is not to document your own 
process of discovery -- save that for an exploration notebook -- but instead it is 
to bring a reader that is new to the data to a full understanding of the important 
features of your data as quickly as possible. In order to do this, your EDA should 
include a detailed, orderly narrative description of what you want your reader to 
know. Do not include any output -- tables, plots, or statistics -- that you do not 
intend to write about.

```{r, data-distribution check}
data$zero_tolerance_law %>% table(useNA = "ifany") %>% as.data.frame()
data$graduated_drivers_license_law %>% table(useNA = "ifany") %>% as.data.frame()
data$per_se_laws %>% table(useNA = "ifany") %>% as.data.frame()
data$minimum_drinking_age %>% table(useNA = "ifany") %>% as.data.frame()
```

```{r, adding binary variuables}

# Binary variable zero_tolerance_law_refined based on zero_tolerance_law
data <- data %>% 
     mutate(zero_tolerance_law_refined = ifelse(zero_tolerance_law < 0.5, 0, 1))

# Binary variable graduated_drivers_license_law_refined based on graduated_drivers_license_law
data <- data %>% 
     mutate(graduated_drivers_license_law_refined = ifelse(graduated_drivers_license_law < 0.5, 0, 1))

# Binary variable per_se_laws_refined based on per_se_laws
data <- data %>% 
     mutate(per_se_laws_refined = ifelse(per_se_laws < 0.5, 0, 1))

# Rounding min drinking age
data <- data %>%
  mutate(
    minimum_drinking_age = round(minimum_drinking_age, 0)
  )
```

```{r, state wise avg fatality rate plot}
data %>% ggplot(aes(reorder(state,desc(total_fatalities_rate)),total_fatalities_rate,
             fill=state)) +
  geom_boxplot(alpha=0.4) +
  # theme_economist_white(gray_bg=F) +
  theme(legend.position="none",axis.text.y=element_text(size=6)) +
  scale_y_continuous() +
  xlab("State") +
  ylab("Total_Fatalities_Rate")
```
```{r}
data %>%
    ggplot(aes(year,total_fatalities_rate,color=state)) +
    geom_point(alpha=0.4) +
    geom_smooth(method="lm") +
    facet_wrap(~state,scales="free_y") +
    theme_economist_white(gray_bg=F) +
    theme(legend.position="none",axis.text.x=element_text(angle=45,hjust=1,vjust=1,size=6),
          axis.text.y=element_text(size=6)) +
    theme(strip.text=element_text(size=4)) +
    scale_y_continuous() +
    xlab("State") +
    ylab("Total_Fatalities_Rate")
```
```{r, histogram plot}
h1 = ggplot(data, aes(x = total_traffic_fatalities)) + 
               scale_x_continuous(name = "Total Fatal.") +
               plot_hist + plot_theme

h2 = ggplot(data, aes(x = log(total_traffic_fatalities))) + 
               scale_x_continuous(name = "Log(Total Fatal.)") +
               plot_hist + plot_theme + plot_theme

h3 = ggplot(data, aes(x = total_weekend_fatalities)) +
               scale_x_continuous(name = "Wknd. Fatal.") +
               plot_hist + plot_theme

h4 = ggplot(data, aes(x = log(total_weekend_fatalities))) +
               scale_x_continuous(name = "Log(Wknd. Fatal.)") +
               plot_hist + plot_theme + plot_theme

h5 = ggplot(data, aes(x = total_nighttime_fatalities)) +
               scale_x_continuous(name = "Night Fatal.") +
               plot_hist + plot_theme

h6 = ggplot(data, aes(x = log(total_nighttime_fatalities))) +
               scale_x_continuous(name = "Log(Night Fatal.)") +
               plot_hist + plot_theme + plot_theme

h7 = ggplot(data, aes(x = total_fatalities_per_100_million_miles)) +
               scale_x_continuous(name = "Total per Mile Rate") +
               plot_hist + plot_theme

h8 = ggplot(data, aes(x = log(total_fatalities_per_100_million_miles))) +
               scale_x_continuous(name = "Log(Total per Mile Rate)") +
               plot_hist + plot_theme + plot_theme

h9 = ggplot(data, aes(x = weekend_fatalities_per_100_million_miles)) +
               scale_x_continuous(name = "Wknd per Mile Rate") +
               plot_hist + plot_theme

h10 = ggplot(data, aes(x = log(weekend_fatalities_per_100_million_miles))) +
               scale_x_continuous(name = "Log(Wknd per Mile Rate)") +
               plot_hist + plot_theme + plot_theme

h11 = ggplot(data, aes(x = nighttime_fatalities_per_100_million_miles)) +
               scale_x_continuous(name = "Night per Mile Rate") +
               plot_hist + plot_theme

h12 = ggplot(data, aes(x = log(nighttime_fatalities_per_100_million_miles))) +
               scale_x_continuous(name = "Log(Night per Mile Rate)") +
               plot_hist + plot_theme + plot_theme

h13 = ggplot(data, aes(x = total_fatalities_rate)) +
               scale_x_continuous(name = "Total per Cap Rate") +
               plot_hist + plot_theme

h14 = ggplot(data, aes(x = log(total_fatalities_rate))) +
               scale_x_continuous(name = "Log(Total per Cap Rate)") +
               plot_hist + plot_theme + plot_theme

h15 = ggplot(data, aes(x = weekend_fatalities_rate)) +
               scale_x_continuous(name = "Wknd per Cap Rate") +
               plot_hist + plot_theme

h16 = ggplot(data, aes(x = log(weekend_fatalities_rate))) +
               scale_x_continuous(name = "Log(Wknd per Cap Rate)") +
               plot_hist + plot_theme + plot_theme

h17 = ggplot(data, aes(x = nighttime_fatalities_rate)) +
               scale_x_continuous(name = "Night per Cap Rate") +
               plot_hist + plot_theme

h18 = ggplot(data, aes(x = log(nighttime_fatalities_rate))) +
               scale_x_continuous(name = "Log(Night per Cap Rate)") +
               plot_hist + plot_theme

grid.arrange(h1,h2,h3,h4,h5,h6,h7,h8,h9,h10,h11,h12,h13,h14,h15,h16,h17,h18,
             nrow = 5, ncol = 4,
             top=quote("Fatalities, Log(Fatalities) Distribution"))
```
```{r, state box plot}
box_st_1 <- data %>%
  group_by(state) %>%
  ggplot(aes(x = reorder(state,total_traffic_fatalities), 
             y = total_traffic_fatalities)) +
  geom_boxplot() +
  labs(x = "States",  y = "Total Fatality rate")

box_st_2 <- data %>%
  group_by(state) %>%
  ggplot(aes(x = reorder(state,total_weekend_fatalities), 
             y = total_weekend_fatalities)) +
  geom_boxplot() +
  labs(x = "States",  y = "Wknd. Fatality rate")

box_st_3 <- data %>%
  group_by(state) %>%
  ggplot(aes(x = reorder(state,total_nighttime_fatalities), 
             y = total_nighttime_fatalities)) +
  geom_boxplot() +
  labs(x = "States",  y = "Night Fatality rate")

box_st_4 <- data %>%
  group_by(state) %>%
  ggplot(aes(x = reorder(state,total_fatalities_per_100_million_miles), 
             y = total_fatalities_per_100_million_miles)) +
  geom_boxplot() +
  labs(x = "States",  y = "Total per Mile Rate")

box_st_5 <- data %>%
  group_by(state) %>%
  ggplot(aes(x = reorder(state,weekend_fatalities_per_100_million_miles), 
             y = weekend_fatalities_per_100_million_miles)) +
  geom_boxplot() +
  labs(x = "States",  y = "Wknd per Mile Rate")

box_st_6 <- data %>%
  group_by(state) %>%
  ggplot(aes(x = reorder(state,nighttime_fatalities_per_100_million_miles), 
             y = nighttime_fatalities_per_100_million_miles)) +
  geom_boxplot() +
  labs(x = "States",  y = "Night per Mile Rate")

box_st_7 <- data %>%
  group_by(state) %>%
  ggplot(aes(x = reorder(state,total_fatalities_rate), 
             y = total_fatalities_rate)) +
  geom_boxplot() +
  labs(x = "States",  y = "Total per Cap. Rate")

box_st_8 <- data %>%
  group_by(state) %>%
  ggplot(aes(x = reorder(state,weekend_fatalities_rate), 
             y = weekend_fatalities_rate)) +
  geom_boxplot() +
  labs(x = "States",  y = "Wknd. per Cap. Rate")

box_st_9 <- data %>%
  group_by(state) %>%
  ggplot(aes(x = reorder(state,nighttime_fatalities_rate), 
             y = nighttime_fatalities_rate)) +
  geom_boxplot() +
  labs(x = "States",  y = "Night per Cap Rate")

grid.arrange(box_st_1,box_st_2,box_st_3,box_st_4,box_st_5,box_st_6,box_st_7,
             box_st_8,box_st_9,nrow = 3, ncol = 3,
             top=quote("Boxplot of Fatalities by State 1980 - 2004"))
```
```{r, year box plot}
box_yr_1 <- data %>%
  group_by(year) %>%
  ggplot(aes(x = year_of_observation, 
             y = total_traffic_fatalities)) +
  geom_boxplot() +
  labs(x = "years",  y = "Total Fatality rate")

box_yr_2 <- data %>%
  group_by(year) %>%
  ggplot(aes(x = year_of_observation, 
             y = total_weekend_fatalities)) +
  geom_boxplot() +
  labs(x = "years",  y = "Wknd. Fatality rate")

box_yr_3 <- data %>%
  group_by(year) %>%
  ggplot(aes(x = year_of_observation, 
             y = total_nighttime_fatalities)) +
  geom_boxplot() +
  labs(x = "years",  y = "Night Fatality rate")

box_yr_4 <- data %>%
  group_by(year) %>%
  ggplot(aes(x = year_of_observation, 
             y = total_fatalities_per_100_million_miles)) +
  geom_boxplot() +
  labs(x = "years",  y = "Total per Mile Rate")

box_yr_5 <- data %>%
  group_by(year) %>%
  ggplot(aes(x = year_of_observation, 
             y = weekend_fatalities_per_100_million_miles)) +
  geom_boxplot() +
  labs(x = "years",  y = "Wknd per Mile Rate")

box_yr_6 <- data %>%
  group_by(year) %>%
  ggplot(aes(x = year_of_observation, 
             y = nighttime_fatalities_per_100_million_miles)) +
  geom_boxplot() +
  labs(x = "years",  y = "Night per Mile Rate")

box_yr_7 <- data %>%
  group_by(year) %>%
  ggplot(aes(x = year_of_observation, 
             y = total_fatalities_rate)) +
  geom_boxplot() +
  labs(x = "years",  y = "Total per Cap. Rate")

box_yr_8 <- data %>%
  group_by(year) %>%
  ggplot(aes(x = year_of_observation, 
             y = weekend_fatalities_rate)) +
  geom_boxplot() +
  labs(x = "years",  y = "Wknd. per Cap. Rate")

box_yr_9 <- data %>%
  group_by(year) %>%
  ggplot(aes(x = year_of_observation, 
             y = nighttime_fatalities_rate)) +
  geom_boxplot() +
  labs(x = "years",  y = "Night per Cap Rate")

grid.arrange(box_yr_1,box_yr_2,box_yr_3,box_yr_4,box_yr_5,box_yr_6,box_yr_7,
             box_yr_8,box_yr_9, nrow = 3, ncol = 3,
             top=quote("Boxplot of Fatalities over years 1980 - 2004"))
```
```{r, histogram of explanatory variables}
ex_h1 = ggplot(data, aes(x = minimum_drinking_age)) + 
               scale_x_continuous(name = "Minimum Age") +
               plot_hist + plot_theme

ex_h2 = ggplot(data, aes(x = log(minimum_drinking_age))) + 
               scale_x_continuous(name = "Log(Minimum Age)") +
               plot_hist + plot_theme

ex_h3 = ggplot(data, aes(x = state_population)) + 
               scale_x_continuous(name = "State Population") +
               plot_hist + plot_theme

ex_h4 = ggplot(data, aes(x = log(state_population))) + 
               scale_x_continuous(name = "Log(State Population)") +
               plot_hist + plot_theme

ex_h5 = ggplot(data, aes(x = vehicle_miles)) + 
               scale_x_continuous(name = "Miles Trvld") +
               plot_hist + plot_theme

ex_h6 = ggplot(data, aes(x = log(vehicle_miles))) + 
               scale_x_continuous(name = "Log(Miles Trvld)") +
               plot_hist + plot_theme

ex_h7 = ggplot(data, aes(x = unemployment_rate)) + 
               scale_x_continuous(name = "Unemp. Rate") +
               plot_hist + plot_theme

ex_h8 = ggplot(data, aes(x = log(unemployment_rate))) + 
               scale_x_continuous(name = "Log(Unemp. Rate)") +
               plot_hist + plot_theme

ex_h9 = ggplot(data, aes(x = pct_population_14_to_24)) + 
               scale_x_continuous(name = "% Pop 14-24") +
               plot_hist + plot_theme

ex_h10 = ggplot(data, aes(x = log(pct_population_14_to_24))) + 
               scale_x_continuous(name = "Log(% Pop 14-24)") +
               plot_hist + plot_theme

ex_h11 = ggplot(data, aes(x = vehicle_miles_per_capita)) + 
               scale_x_continuous(name = "Miles Driven") +
               plot_hist + plot_theme

ex_h12 = ggplot(data, aes(x = log(vehicle_miles_per_capita))) + 
               scale_x_continuous(name = "log(Miles Driven)") +
               plot_hist + plot_theme

ex_h13 = ggplot(data, aes(x = as.numeric(speed_limit))) + 
               scale_x_continuous(name = "Speed Limit") +
               plot_hist + plot_theme

ex_h14 = ggplot(data, aes(x = log(as.numeric(speed_limit)))) + 
               scale_x_continuous(name = "Log(Speed Limit)") +
               plot_hist + plot_theme

grid.arrange(ex_h1,ex_h2,ex_h3,ex_h4,ex_h5,ex_h6,ex_h7,
             ex_h8,ex_h9,ex_h10,ex_h11,ex_h12,ex_h13,ex_h14,
             nrow = 7, ncol = 2,
             top=quote("Distributions of Explanatory Variables and Logs"))
```

```{r, state level avg}
state.mean <- aggregate(data[, c("total_fatalities_rate", 
                               "state_population",
                               "unemployment_rate",
                               "pct_population_14_to_24",
                               "vehicle_miles",
                               "vehicle_miles_per_capita")], 
                         data["state"], FUN = mean)
ggpairs(state.mean, size = 5, cardinality_threshold = 60) + 
  ggtitle(label = "Year Averages over State")
```
```{r, year level avg}
year.mean <- aggregate(data[, c("total_fatalities_rate", 
                                    "state_population",
                                    "unemployment_rate",
                                    "pct_population_14_to_24",
                                    "vehicle_miles",
                                    "vehicle_miles_per_capita")], 
                         data["year_of_observation"], FUN = mean)
ggpairs(year.mean, size = 5, cardinality_threshold=60) + 
  ggtitle(label = "State Averages over time (years)")
```
```{r, fatality rate vs year}
yearly.law.mean <- aggregate(data[,c("total_fatalities_rate", 
                                        "minimum_drinking_age", 
                                        "zero_tolerance_law",
                                        "speed_limit_70plus",
                                        "graduated_drivers_license_law",
                                        "per_se_laws",
                                        "blood_alcohol_limit",
                                        "primary_seatbelt_law",
                                        "secondary_seatbelt_law")], 
                              data["year_of_observation"], FUN = mean)
ggpairs(yearly.law.mean, size = 5, cardinality_threshold=100) + 
  ggtitle(label = "State Law Proportions over time (years)")
```
```{r}
state.law.mean <- aggregate(data[,c("total_fatalities_rate", 
                                        "minimum_drinking_age", 
                                        "zero_tolerance_law",
                                        "speed_limit_70plus",
                                        "graduated_drivers_license_law",
                                        "per_se_laws",
                                        "blood_alcohol_limit",
                                        "primary_seatbelt_law",
                                        "secondary_seatbelt_law")], 
                              data["state"], FUN = mean)
ggpairs(state.law.mean, size = 5, cardinality_threshold=100) + 
  ggtitle(label = "State Law Time Proportions vs State")
```

# (15 points) Preliminary Model

Estimate a linear regression model of *totfatrte* on a set of dummy variables for 
the years 1981 through 2004 and interpret what you observe. In this section, you 
should address the following tasks: 

- Why is fitting a linear model a sensible starting place? 
- What does this model explain, and what do you find in this model? 
- Did driving become safer over this period? Please provide a detailed explanation.
- What, if any, are the limitation of this model. In answering this, please consider 
**at least**: 
    - Are the parameter estimates reliable, unbiased estimates of the truth? Or, 
    are they biased due to the way that the data is structured?
    - Are the uncertainty estimate reliable, unbiased estimates of sampling based 
    variability? Or, are they biased due to the way that the data is structured? 

```{r, base model}
model.base <- lm(total_fatalities_rate ~ year_of_observation, data=data)
summary(model.base)
```

# (15 points) Expanded Model 

Expand the **Preliminary Model** by adding variables related to the following 
concepts: 

- Blood alcohol levels 
- Per se laws
- Primary seat belt laws (Note that if a law was enacted sometime within a year the 
fraction of the year is recorded in place of the zero-one indicator.)
- Secondary seat belt laws 
- Speed limits faster than 70 
- Graduated drivers licenses 
- Percent of the population between 14 and 24 years old
- Unemployment rate
- Vehicle miles driven per capita. 

```{r Expanded Model}
model.expanded.1 = lm(total_fatalities_rate ~ year_of_observation + 
                                              blood_alcohol_limit +
                                              per_se_laws + 
                                              primary_seatbelt_law + 
                                              secondary_seatbelt_law + 
                                              speed_limit_70plus + 
                                              graduated_drivers_license_law +               
                                              log(pct_population_14_to_24) + 
                                              log(unemployment_rate) + 
                                              log(vehicle_miles_per_capita), 
                data = data) 
summary(model.expanded.1)
```

If it is appropriate, include transformations of these variables. Please carefully 
explain carefully your rationale, which should be based on your EDA, behind any 
transformation you made. If no transformation is made, explain why transformation 
is not needed. 

- How are the blood alcohol variables defined? Interpret the coefficients that you 
estimate for this concept. 
- Do *per se laws* have a negative effect on the fatality rate? 
- Does having a primary seat belt law? 

```{r}
shapiro.test(log(data$pct_population_14_to_24))
shapiro.test(data$pct_population_14_to_24)

shapiro.test(log(data$unemployment_rate))
shapiro.test(data$unemployment_rate)

shapiro.test(log(data$vehicle_miles_per_capita))
shapiro.test(data$vehicle_miles_per_capita)
```
# (15 points) State-Level Fixed Effects 

Re-estimate the **Expanded Model** using fixed effects at the state level. 

- What do you estimate for coefficients on the blood alcohol variables? How do 
the coefficients on the blood alcohol variables change, if at all? 
- What do you estimate for coefficients on per se laws? How do the coefficients 
on per se laws change, if at all? 
- What do you estimate for coefficients on primary seat-belt laws? How do the 
coefficients on primary seatbelt laws change, if at all? 

Which set of estimates do you think is more reliable? Why do you think this? 

- What assumptions are needed in each of these models?  
- Are these assumptions reasonable in the current context?

```{r Fixed Effects models}
data.new <- data
row.names(data.new) <- NULL
pool.model <- plm(total_fatalities_rate ~ year_of_observation + 
                                              blood_alcohol_limit +
                                              per_se_laws + 
                                              primary_seatbelt_law + 
                                              secondary_seatbelt_law + 
                                              speed_limit_70plus + 
                                              graduated_drivers_license_law +               
                                              log(pct_population_14_to_24) + 
                                              log(unemployment_rate) + 
                                              log(vehicle_miles_per_capita), 
                data = data.new, index=c("state", "year_of_observation"), model="pooling")

fd.model <- plm(total_fatalities_rate ~ year_of_observation + 
                                              blood_alcohol_limit +
                                              per_se_laws + 
                                              primary_seatbelt_law + 
                                              secondary_seatbelt_law + 
                                              speed_limit_70plus + 
                                              graduated_drivers_license_law +               
                                              log(pct_population_14_to_24) + 
                                              log(unemployment_rate) + 
                                              log(vehicle_miles_per_capita), 
                data = data.new, index=c("state", "year_of_observation"), model="fd")

between.model <- plm(total_fatalities_rate ~ year_of_observation + 
                                              blood_alcohol_limit +
                                              per_se_laws + 
                                              primary_seatbelt_law + 
                                              secondary_seatbelt_law + 
                                              speed_limit_70plus + 
                                              graduated_drivers_license_law +               
                                              log(pct_population_14_to_24) + 
                                              log(unemployment_rate) + 
                                              log(vehicle_miles_per_capita), 
                data = data.new, index=c("state", "year_of_observation"), model="between")

within.model <- plm(total_fatalities_rate ~ year_of_observation + 
                                              blood_alcohol_limit +
                                              per_se_laws + 
                                              primary_seatbelt_law + 
                                              secondary_seatbelt_law + 
                                              speed_limit_70plus + 
                                              graduated_drivers_license_law +               
                                              log(pct_population_14_to_24) + 
                                              log(unemployment_rate) + 
                                              log(vehicle_miles_per_capita), 
                data = data.new, index=c("state", "year_of_observation"), model="within")
summary(pool.model)
summary(fd.model)
summary(between.model)
summary(within.model)

stargazer(pool.model, fd.model, between.model, within.model, type = "text",
          omit.stat = c("ser","f","adj.rsq"), dep.var.labels = "",
          column.labels = c("Pooled", "FD", "Between", "Within"))

pwfdtest(fd.model, data=dat, index=c("state","year"), h0="fe")
pwfdtest(fd.model, data=dat, index=c("state","year"), h0="fd")
pFtest(within.model, pool.model)
```

# (10 points) Consider a Random Effects Model 

Instead of estimating a fixed effects model, should you have estimated a random 
effects model?

- Please state the assumptions of a random effects model, and evaluate whether 
these assumptions are met in the data. 
- If the assumptions are, in fact, met in the data, then estimate a random effects 
model and interpret the coefficients of this model. Comment on how, if at all, 
the estimates from this model have changed compared to the fixed effects model. 
- If the assumptions are **not** met, then do not estimate the data. But, also 
comment on what the consequences would be if you were to *inappropriately* 
estimate a random effects model. Would your coefficient estimates be biased or not? 
Would your standard error estimates be biased or not? Or, would there be some other 
problem that might arise?

# (10 points) Model Forecasts 

The COVID-19 pandemic dramatically changed patterns of driving. Find data 
(and include this data in your analysis, here) that includes some measure of 
vehicle miles driven in the US. Your data should at least cover the period from 
January 2018 to as current as possible. With this data, produce the following 
statements: 

- Comparing monthly miles driven in 2018 to the same months during the pandemic: 
  - What month demonstrated the largest decrease in driving? How much, in percentage 
  terms, lower was this driving? 
  - What month demonstrated the largest increase in driving? How much, in percentage 
  terms, higher was this driving? 
  
Now, use these changes in driving to make forecasts from your models. 

- Suppose that the number of miles driven per capita, increased by as much as the 
COVID boom. Using the FE estimates, what would the consequences be on the number 
of traffic fatalities? Please interpret the estimate.
- Suppose that the number of miles driven per capita, decreased by as much as the 
COVID bust. Using the FE estimates, what would the consequences be on the number 
of traffic fatalities? Please interpret the estimate.

# (5 points) Evaluate Error 

If there were serial correlation or heteroskedasticity in the idiosyncratic errors 
of the model, what would be the consequences on the estimators and their standard 
errors? Is there any serial correlation or heteroskedasticity? 