
---
title: 'Lab 3: Panel Models'
subtitle: 'US Traffic Fatalities: 1980 - 2004'
output: 
  bookdown::pdf_document2: default
---

```{r load packages, echo=FALSE, message=FALSE}
rm(list = ls())
library(tidyverse)
library(knitr)
library(data.table)
library(plm)
library(ggrepel)
library(ggplot2)
library(stargazer)
```

# U.S. traffic fatalities: 1980-2004

<https://github.com/srilamaiti/w271_summer_2023_elizabeth_emily_michael_olivia_srila_lab3>

In this lab, we are asking you to answer the following **causal** question:

> **"Do changes in traffic laws affect traffic fatalities?"**

To answer this question, please complete the tasks specified below using the data provided in `data/driving.Rdata`. This data includes 25 years of data that cover changes in various state drunk driving, seat belt, and speed limit laws.

Specifically, this data set contains data for the 48 continental U.S. states from 1980 through 2004. Various driving laws are indicated in the data set, such as the alcohol level at which drivers are considered legally intoxicated. There are also indicators for "per se" laws---where licenses can be revoked without a trial---and seat belt laws. A few economics and demographic variables are also included. The description of the each of the variables in the dataset is also provided in the dataset.

```{r load data, echo = TRUE}
load(file="~/MIDS/Summer23/w271_summer_2023_elizabeth_emily_michael_olivia_srila_lab3/data/driving.RData")

## please comment these calls in your work 
glimpse(data)
desc

unique(data$state)
```

# (30 points, total) Build and Describe the Data

1.  (5 points) Load the data and produce useful features. Specifically:
    -   Produce a new variable, called `speed_limit` that re-encodes the data that is in `sl55`, `sl65`, `sl70`, `sl75`, and `slnone`;
    -   Produce a new variable, called `year_of_observation` that re-encodes the data that is in `d80`, `d81`, ... , `d04`.
    -   Produce a new variable for each of the other variables that are one-hot encoded (i.e. `bac*` variable series).
    -   Rename these variables to sensible names that are legible to a reader of your analysis. For example, the dependent variable as provided is called, `totfatrte`. Pick something more sensible, like, `total_fatalities_rate`. There are few enough of these variables to change, that you should change them for all the variables in the data. (You will thank yourself later.)

```{r}
data <- data %>% 
  mutate(year_of_observation =
           case_when(
             data$d80 == 1 ~ 1980,
             data$d81 == 1 ~ 1981,
             data$d82 == 1 ~ 1982,
             data$d83 == 1 ~ 1983,
             data$d84 == 1 ~ 1984,
             data$d85 == 1 ~ 1985,
             data$d86 == 1 ~ 1986,
             data$d87 == 1 ~ 1987,
             data$d88 == 1 ~ 1988,
             data$d89 == 1 ~ 1989,
             data$d90 == 1 ~ 1990,
             data$d91 == 1 ~ 1991,
             data$d92 == 1 ~ 1992,
             data$d93 == 1 ~ 1993,
             data$d94 == 1 ~ 1994,
             data$d95 == 1 ~ 1995,
             data$d96 == 1 ~ 1996,
             data$d97 == 1 ~ 1997,
             data$d98 == 1 ~ 1998,
             data$d99 == 1 ~ 1999,
             data$d00 == 1 ~ 2000,
             data$d01 == 1 ~ 2001,
             data$d02 == 1 ~ 2002,
             data$d03 == 1 ~ 2003,
             data$d04 == 1 ~ 2004,
             TRUE ~ 00000
           )
)
data %>% filter(year_of_observation == 00000)   

data <- data %>%
  mutate(
    blood_alc =
      case_when(
        bac10 > .5 ~ .1,
        bac08 > .5 ~ .08,
        bac08 == 0 & bac10 == 0 ~ 0,
        bac10 > bac08 ~ .1,
        TRUE ~ .08
      )
  )
data <- data %>%
  mutate(speed_limit =
           case_when(
             data$sl55 > .5 ~ 55,
             data$sl65 > .5 ~ 65,
             data$sl70 > .5 ~ 70,
             data$sl75 > .5 ~ 75,
             data$slnone > .5 ~ 0,
             data$year == 1996 & data$state == 6 ~ 75,
             data$year == 1996 & data$state == 11 ~ 70,
             data$year == 1995 & data$state == 21 ~ 65,
             data$year == 1997 & data$state == 24 ~ 70,
             data$year == 1988 & data$state == 47 ~ 65
           ))


old_names <- c("seatbelt", "minage", "zerotol", "gdl", "perse", "totfat",
               "nghtfat", "wkndfat", "totfatpvm", "nghtfatpvm", "wkndfatpvm",
               "statepop", "totfatrte", "nghtfatrte", "wkndfatrte", "vehicmiles",
               "unem", "perc14_24", "vehicmilespc", "sbprim", "sbsecon")

new_names <- c("seatbelt_law", "min_drink_age", "zero_tol", "grad_driver_license", "per_se",
               "tot_fat", "tot_night_fat", "tot_wknd_fat", "tot_fat_per_mil_miles",
               "tot_nght_fat_per_mil_miles", "tot_wknd_fat_per_mil_miles",
               "state_pop", "total_fatalities_rate", "tot_nght_fat_rate", "tot_wknd_fat_rate",
               "miles_traveled_b", "unemployment", "percent_pop_14_24", "vehic_miles_pc", "primary_seatbelt", "secondary_seatbelt")

data <- data %>%
  setnames(old = old_names, new = new_names)
```

```{r select columns}
data <- data %>%
  dplyr::select("year", "state", "seatbelt_law", "min_drink_age", "zero_tol",
         "grad_driver_license", "per_se", "tot_fat", "tot_night_fat", "tot_wknd_fat",
         "tot_fat_per_mil_miles","tot_nght_fat_per_mil_miles", "tot_wknd_fat_per_mil_miles",
         "state_pop", "total_fatalities_rate", "tot_nght_fat_rate", "tot_wknd_fat_rate",
         "miles_traveled_b", "unemployment", "percent_pop_14_24", "vehic_miles_pc",
         "speed_limit", "year_of_observation", "blood_alc", "primary_seatbelt", "secondary_seatbelt")
# data <- merge(x = data, y = states, by.x = "state", by.y = "row")
# 
# data
```


```{r}
data$state <- as.character(data$state)
data <- data %>% 
  mutate(state = dplyr::recode(state,
    "1" = "AL",
    "3" = "AZ",
    "4" = "AR",
    "5" = "CA",
    "6" = "CO",
    "7" = "CT",
    "8" = "DE",
    "10" = "FL",
    "11" = "GA",
    "13" = "ID",
    "14" = "IL",
    "15" = "IN",
    "16" = "IA",
    "17" = "KS",
    "18" = "KY",
    "19" = "LA",
    "20" = "ME",
    "21" = "MD",
    "22" = "MA",
    "23" = "MI",
    "24" = "MN",
    "25" = "MS",
    "26" = "MO",
    "27" = "MT",
    "28" = "NE",
    "29" = "NV",
    "30" = "NH",
    "31" = "NJ",
    "32" = "NM",
    "33" = "NY",
    "34" = "NC",
    "35" = "ND",
    "36" = "OH",
    "37" = "OK",
    "38" = "OR",
    "39" = "PA",
    "40" = "RI",
    "41" = "SC",
    "42" = "SD",
    "43" = "TN",
    "44" = "TX",
    "45" = "UT",
    "46" = "VT",
    "47" = "VA",
    "48" = "WA",
    "49" = "WV",
    "50" = "WI",
    "51" = "WY"
    ))

```


# ((15 points) Preliminary Model (Elizabeth)
Estimate a linear regression model of *totfatrte* on a set of dummy variables for the years 1981 through 2004 and interpret what you observe. In this section, you should address the following tasks:

-   Why is fitting a linear model a sensible starting place?
-   What does this model explain, and what do you find in this model?
-   Did driving become safer over this period? Please provide a detailed explanation.
-   What, if any, are the limitation of this model. In answering this, please consider **at least**:
    -   Are the parameter estimates reliable, unbiased estimates of the truth? Or, are they biased due to the way that the data is structured?
    -   Are the uncertainty estimate reliable, unbiased estimates of sampling based variability? Or, are they biased due to the way that the data is structured?

A linear model is a sensible starting place because we have non-independence in this data set. This dataset also contains some variability, since we have repeated measures taken over time, (i.e, finding the fatality rate of each state every year). A linear model can help us determine the relationships between different variables and the outcome and allow us to better understand the source of variability within the dataset.

```{r}
lsdv_mod <- plm(total_fatalities_rate ~  min_drink_age + state, index=c("state", "year"), model = "within",data=data)
summary(lsdv_mod)
```

We incorporate dummy variables for the year and for all states. The model doesn't include any other variable, so omitted variable bias could be present, however, the results imply that minimum drinking age does have an impact on the total fatality rate within states.


# (15 points) Expanded Model (Emily)


```{r}
# create speed limit over 70 variable
data = data %>% mutate(speed_lim_70 = ifelse(speed_limit == 55 | speed_limit == 65, 0, 1))

```

```{r checking normality}
# testing variables for normality
shap_test <- c("total_fatalities_rate", "percent_pop_14_24", "unemployment", "vehic_miles_pc")

shapiro_test_results <- lapply(shap_test, function(var_name) {
  if (var_name != "state" && var_name != "year") {  # Exclude 'state' and 'year' from the test
    shapiro_test <- shapiro.test(data[[var_name]])
    return(list(Variable = var_name, p_value = shapiro_test$p.value))
  }
})

shapiro_test_df <- do.call(rbind, shapiro_test_results)

print(shapiro_test_df)

```
We used the Shapiro-Wilks test to test for normality with the total fatalities rate, percent of population 14-24, unemployment, and vehicle miles per capita variables. The test resulted in very small p-values for each, so we decided to log transform these variables in our expanded model.

```{r}
exp_mod <- plm(log(total_fatalities_rate) ~ year + min_drink_age + blood_alc + per_se +
           primary_seatbelt + secondary_seatbelt + speed_lim_70 + grad_driver_license +
           log(percent_pop_14_24) + log(unemployment) + log(vehic_miles_pc),
           index=c("state", "year"), model = "pooling", data=data)

summary(exp_mod)
```
Based on the expanded model, the vehicle miles driven per capita, unemployment rate, and percent of population between 14 and 24 were all highly statistically significant. Both per se laws and primary seat belt laws had negative effects on fatality rate. Additionally, per se laws and minimum drinking age laws had less significant but still negative effects on fatality rates. 

The blood alcohol content (BAC) variable represents the alcohol level at which drivers are considered legally intoxicated. BAC is calculated by Alcohol consumed in grams divided by body weight in grams, then multiplied by a constant that represents biological sex. This number is then multiplied by 100 to give a final measurement of grams per 100mL of blood. The coefficient estimate for the blood alcohol content (BAC) variable in our expanded model is -0.0042982. The negative relationship suggests that stricter blood alcohol content laws may be associated with a decrease in the log-transformed total fatalities rate. However, this impact was not found to be statistically significant (p-value = 0.982275). 



